Approximation as Differentiation

### 2025-06-08 main/first-order.tex
Lifting.

The action of the families construction, regarded as a covariant 2-functor $\Cat \to \Cat$, sends functors $F:
\cat{C} \to \cat{D}$ to functors $\Fam(F): \Fam(\cat{C}) \to \Fam(\cat{D})$ that use $F$ to reassign the
target category, sending objects $(X, \partial X)$ to $(X, F \comp \partial X)$ and morphisms $(f, \partial
f)$ to $(f, F \comp \partial f)$, where $F \comp \partial f$ denotes the natural transformation with $(F \comp
\partial f)(x) = F(\partial f(x))$. If $F$ happens to be a (strong) monad on $\cat{C}$, then $\Fam(F)$ is a
(strong) monad on $\Fam(\cat{C})$, defined by lifting the original monad $F: \cat{C} \to \cat{C}$ pointwise
over families.

@@

The total category $\Fam(\LatGal)$ is then a suitable setting for interpreting first-order programs for \GPS.
It has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every $x \in X$, a bounded lattice
$\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of a
function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial f(x): \partial X(x) \to \partial
Y(f(x))$. This is precisely the setup we identified informally at the beginning of this section. The
reindexing along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial
g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois connection at each
point, using the baseline (unapproximated) function $f$.

@@

For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
\RR^n$ at a point $x$ have the following form:

>>>>

The action of the families construction, regarded as a covariant 2-functor $\Cat \to \Cat$, sends functors $F:
\cat{C} \to \cat{D}$ to functors $\Fam(F): \Fam(\cat{C}) \to \Fam(\cat{D})$ that use $F$ to reassign the
target category by postcomposing with $F$, sending objects $(X, \partial X)$ to $(X, F \comp \partial X)$ and
morphisms $(f, \partial f)$ to $(f, F \comp \partial f)$, where $F \comp \partial f$ denotes the natural
transformation with $(F \comp \partial f)(x) = F(\partial f(x))$. If $F$ happens to be a (strong) monad on
$\cat{C}$, then $\Fam(F)$ is a (strong) monad on $\Fam(\cat{C})$, computed by lifting the original monad $F:
\cat{C} \to \cat{C}$ pointwise over families.

@@

The total category $\Fam(\LatGal)$ is then a suitable setting for interpreting first-order programs for \GPS.
It has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every $x \in X$, a bounded lattice
$\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of a
function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial f(x): \partial X(x) \to \partial
Y(f(x))$. The reindexing along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f,
\reindex{\partial g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois
connection at each point, using the baseline (unapproximated) function $f$.

For a first-order language with a primitive type $\syn{nat}$ and primitive operations such as $+$, the
``leaves'' of the approximation lattices would provided by a model of $\syn{nat}$ and its operations in
$\Fam(\LatGal)$, for example assigning $\syn{nat}$ to the object $(\Nat,n \mapsto \Two)$, where the family is
fully present ($\top$) or fully absent ($\bot$) from the approximation (as shown in the example in component
is constantly $\Two = (\bot,\top)$, the two-point lattice indicating whether a given natural number
\secref{introduction}). These ``leaf'' objects would then be combined by the interpretation of composite types
like sums and products, providing suitable derived approximation structures for composite values.

One might also like to allow the programmer be able to embed additional ``approximation points'' into
composite structures, by providing the language with a \emph{lifting} monad $\syn{L}$. Rather than the usual
definition of $\syn{list}\;\syn{nat}$ as $\mu\alpha.1 + \syn{nat} \times \alpha$, one could then define a type
of ``approximation lists'' $\mu\alpha.\syn{L}\;(1 + \syn{L}\;\syn{nat} \times \alpha)$, along with an
``approximating'' version of multiplication $*: \syn{L}\;\syn{nat} \times \syn{L}\;\syn{nat} \to
\syn{L}\;\syn{nat}$.

@@

For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
\RR^n$ at a point $x$ have the following form \todo{Can probably delete now as covered in
\secref{approx-as-tangents}}:
